On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-10-25 manuscript/pnas_revision.tex
everything but the simulation
manuscript has lowdimviz figure now
add last figure
begun section

  Let $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that

@@

\section{Application: global spread of SARS-CoV-2}
\label{sec:sars:cov:app}

>>>>

  Let symmetric $\QQ \in \mathcal{S}(d)$ be non-positive, i.e.,  suppose that

@@

\section{Empirical Studies}

\begin{figure}[!ht]
	\centering
\includegraphics[width=\linewidth]{lowDimVis.pdf}
\vspace{-2em}
\caption{All methods agree.}\label{fig:lowDim}
\end{figure}

\begin{figure*}[!t]
	\centering
	\includegraphics[width=\linewidth]{truthMeanFig.pdf}
	\caption{Frobenius norm errors obtained by first-order approximation $te^{t\QQ}\JJ$ and by affine-corrected first-order approximation
		$te^{t\QQ}\JJ - t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$
		(\cref{thm:bride:of:frankinstein}) under increasingly relaxed
		assumptions.  Within each assumption set, we average over 20
		independent Monte Carlo simulations of random generator matrices
		for each dimension. Plot \textbf{a} corresponds to asymmetric
		generators with off-diagonal elements having independent and
		identically distributed (iid) standard exponential random
		variables that correspond to the sub-exponential distribution
		hypothesis. Plot \textbf{b} drops the identical distribution
		assumption by allowing each row and column of the generator
		matrix to additively contribute its own mean---itself given by a
		standard exponential---to its corresponding exponentially
		distributed entries. Plot \textbf{c} features rate
		matrices with iid Cauchy entries truncated to be
		positive. Empirically, the results of
		\cref{cor:rig:det:rand:wrapup} extend beyond the symmetric,
		iid and sub-exponential hypotheses, suggesting scope of future work.}
	\label{fig:truthMean}
\end{figure*}

Before applying the naive matrix exponential derivative approximation to the phylogeographic analysis of SARS-CoV-2, we carry out a few targeted studies that illustrate: the empirical performance of the approximate derivative and its affine correction \eqref{eq:better:help:on:sale} (\cref{fig:app}); agreement between CTMC generator matrix empirical posterior distributions generated by surrogate-trajectory Hamiltonian Monte Carlo (HMC) algorithms using approximate derivatives and the truth (\cref{fig:lowDim}); and point estimation of generator matrix element values under different sparsity regimes and different CTMC state space dimensionalities $d$ (\cref{fig:truthMean}).

\section{Application: Global Spread of SARS-CoV-2}
\label{sec:sars:cov:app}
